---
title: "Metabolomics of white tea stored up to 11 years"
author: "Josh Morimoto"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: false
    theme: null
    highlight: null
    toc: true
    toc_depth: 2
bibliography: references/references.bib
csl: references/analytical-chemistry.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, 
  fig.align='center',
  cache=TRUE)

library(knitr)
library(kableExtra)
```

Tea is an important subject for me because it was the model crop on which my
PhD thesis was based. At Tufts, we harvested tea grown in different seasons and
elevations to see how its environment changed its flavor and health properties,
all driven by its metabolism. Like many plants, tea uses chemical signals to
communicate with its environment, since it can't get up and walk away like an
animal or even a microbe could. Evolution has driven myriad metabolic pathways
to allow plants to defend themselves. This makes them particularly complex and
challenging subjects of research, slowing studies on how crop qualities will
change under different conditions, and exacerbated with the increasing threat 
and severity of climate change.

In this study, I discovered that two metabolites change strongly with storage
time.  They correspond to the differential features 607.2477_411 and
130.0507_69, named as _m/z_-RT pairs.

For this demo/tutorial, I looked for data collected from _Camellia sinensis_
samples on MetaboLights[@yurekten2023] and found 
[one study in progress by Wei and Li](https://www.ebi.ac.uk/metabolights/editor/MTBLS13336/overview).  
This study aimed to investigate the sensory and chemical metabolites, both
endogenous to the tea and exogenous from microbial contributions, over long-term
storage.[@mtbls13336] We will focus on the metabolomics data herein.

## IDE considerations
I use VSCode with R and arf as the R frontend. Many people use RStudio or Python
with marimo or Jupyter notebooks but this is what works best for me. I also
manage R/Python environments with conda. For more info, follow [Renkun Ken's
guide on writing R in
VScode](https://renkun.me/2019/12/11/writing-r-in-vscode-a-fresh-start/).

You'll also notice I don't really stick to the R style guidelines. I like my
`=` signs more than typing out `<-` everytime because I learned on Python.
It has never not worked for me. Conversely, I also like naming variables with
`.` rather than snake_case or camelCase. It looks better to my eyeballs.

The conda environment yml can be downloaded [here](https://raw.githubusercontent.com/jmorim/white_tea/refs/heads/master/environment.yml).

I'm also including a couple of packages from github, namely my own `LCMStools`
and Jan Stanstrup's `xcmsVis`. These should be installed with `pak` inside the
R environment.

Finally, there's a good chance if you don't know your environment very well,
you'll end up crashing your R session consistently. Anecdotally, I found
reducing the number of cores used for parallel processing leads to less memory
consumption.  I'm on an Intel Core Ultra 5 245K with 14 cores, and 32 GB of DDR5
RAM.  Setting the MulticoreParam to 4 cores worked well for me, but 6 cores was
consistently too much for my memory. There's probably some way to lazy-load
MsExperiment data, but we'll cover that another time.

If you have problems, feel free to reach out at
[morimoto@motoanalytical.com](mailto:morimoto@motoanalytical.com).

## Libraries
```{r}
library(Spectra)
library(MsBackendMetaboLights)
library(purrr)
library(stringr)
library(dplyr)
library(curl)
library(LCMStools)
library(MsExperiment)
library(BiocParallel)
library(xcms)
library(xcmsVis)
library(plotly)
library(htmlwidgets)
library(colorspace)
library(webchem)
library(SummarizedExperiment)
library(ropls)
library(broom)

if(.Platform$OS == 'Windows'){
  register(bpstart(SnowParam(4, progressbar=FALSE)))
} else if(.Platform$OS == 'unix'){
  register(bpstart(MulticoreParam(4, progressbar=FALSE)))
}
```

Github packages
```{r, eval = FALSE}
library(devtools)
pak::pak('jmorim/LCMStools')
pak::pak('jstanstrup/xcmsVis')
```

For `httpgd` with VSCode and arf

```{r, eval=FALSE}
library(httpgd)
hgd()
hgd_browse()
```

## Loading the study data
First we load the study information from the MetaboLights repository using the
MTBLS ID. This is my first time personally using MetaboLights, so I'm getting a
view on how the data is organized.

```{r}
mtbls.repo = 'MTBLS13336'
all.files = mtbls_list_files(mtbls.repo)
all.files
```

Raw data is stored under *FILES* and study metadata in various formats in the
root directory.

Below is the assay information table, also accessible online
[here](https://www.ebi.ac.uk/metabolights/editor/MTBLS13336/assays).  This
contains sample names, data file names, and some experimental parameters.  If
the data were converted, we would also see a different file under the *Derived
Spectral Data File* column, but the authors kept them in Thermo's proprietary
.raw format. Otherwise, using the `Spectra` package and the
`MsBackendMetaboLights` backend, we could have loaded the *.mzML* files directly
for analysis.

```{r}
study.info = read.table(paste0(mtbls_ftp_path(mtbls.repo), grep('^a_', all.files, value = TRUE)),
    sep = '\t', header = TRUE, check.names = FALSE)

# Can use this when MetaboLights is down if already downloaded, which happened in the middle of this
#study.info = read.table(dir('.', pattern='^a\\_'),
#    sep = '\t', header = TRUE, check.names = FALSE)

knitr::kable(study.info |> head()) |> kable_styling(full_width=TRUE) |> scroll_box(width='100%')
study.info[, c('Raw Spectral Data File', 'Derived Spectral Data File')] |> head()
#specs = Spectra::Spectra('MTBLS13336', filePattern = '.raw', source = MsBackendMetaboLights())
```

Instead, we'll download the raw data and convert herein. This will take a while,
as high resolution mass spec data can be several gigabytes per file, so if
you're playing along at home, you can execute this block and go get a coffee.
In this case, they're ~80 MB per file. I've never worked with Orbitrap data so
that's pretty neat compared to Q-TOF profile data.

Note that I had to replace the `FILES/` prefix from the repo because I don't
need that in my local copy.

I thought parallelizing this with walk would make it go fast but it does not.
Probably faster to download a zip and unzip it locally.

```{r, eval=FALSE, progress=FALSE}
paste0(mtbls_ftp_path(mtbls.repo), mtbls.repo, '.zip')
dir.create('_raw', showWarnings=FALSE)
download.file(
  url = paste0(mtbls_ftp_path(mtbls.repo), mtbls.repo, '.zip'),
  destfile = paste0(mtbls.repo, '.zip')
)
```

That didn't work. MetaboLights doesn't zip raw data in this repo.
The `curl` package has a function `multi_download` that might work faster.

```{r, eval=FALSE, progress=FALSE}
dir.create('_raw', showWarnings=FALSE)

response.table = curl::multi_download(
  urls = paste0(
    mtbls_ftp_path(mtbls.repo), 
    study.info[, 'Raw Spectral Data File']), 
  destfiles = paste0(
    './_raw/',
    stringr::str_replace(study.info[, 'Raw Spectral Data File'], 'FILES/', '')),
  resume = TRUE
)

response.table[,'error'] |> head()
response.table$destfile |> head()
dir('_raw') |> head()
```

This didn't work either due to timeouts with the ftp server. In the end,
downloading the files sequentially with `download_file` was most reliable.

```{r}
dir.create('_raw', showWarnings=FALSE)
if(length(study.info[, 'Raw Spectral Data File']) != length(dir('_raw'))){
  walk(study.info[, 'Raw Spectral Data File'], ~ download.file(
    url = paste0(mtbls_ftp_path(mtbls.repo), .x),
    destfile = paste0('_raw/', stringr::str_replace(.x, 'FILES/', ''))),
    method = 'libcurl'
  )
}
```

## Data conversion
To read files with
`xcms`,[@doi:10.1021/ac051437y;@doi:10.1021/acs.analchem.5c04338] data need to be
converted to *.mzML* format with `MSConvert` from the `ProteoWizard` package.
Since I'm on Linux, I need a docker container to run this with a Windows image.
On Windows, it can run natively. The code below detects the host environment and
chooses the appropriate conversion command. The `convertDtoMZML` function from
my `LCMStools` package was built for Agilent *.d* files, but it should work for
*.raw* files too since the parameters aren't too specific. Docker will prompt
for a password for sudo access. By using the Docker image, you agree to certain
vendor licenses for converting proprietary formats to *.mzML*.

```{r}
raw.files = dir('_raw', pattern = '\\.raw$', full.names = TRUE)

# Check if length mzML files matches length raw files
if(length(raw.files) > length(dir('_mzML', pattern = '\\.mzML$'))){
  if(.Platform$OS == 'Windows'){
    raw.files |> walk(LCMStools::convertDtoMZML, path.out = '_mzML')
  } else if(.Platform$OS == 'unix'){
    walk(raw.files, ~ system(paste0(
      'sudo docker run -it --rm -v $PWD:/data proteowizard/pwiz-skyline-i-agree-to-the-vendor-licenses:latest wine msconvert /data/_raw/', basename(.), ' --filter "peakPicking true 1-" -o /data/_mzML --mzML'
      )
    ))
  }
}

mzml.files = dir('_mzML', pattern = '\\.mzML$', full.names = TRUE)
```

## Initial data inspection
First we'll plot the positive TIC to make sure there aren't any
misbehaving data files or sample runs. In this case, I'm only going to
focus on the positive data for now and leave the negative data as an exercise
for the reader.

```{r}
mzml.files |> head()
pos.files = mzml.files[str_detect(mzml.files, 'POS')]
pos.data = readMsExperiment(pos.files)

plot(chromatogram(pos.data), main = 'Positive TIC')
```

It looks like nothing disastrous happened in the data collection. From the TICs,
it's hard to tell right away if there are any differences since the noise buries
most of the valuable data. Sometimes you can get away with focusing on a certain
region that's obviously different.

Looking closer, the polar region before 100 seconds in retention changes with
intensity. I can infer that these are polar compounds since the column used was
a Waters Acquity bridged-ethylene hybrid (BEH) C18 column, which is reversed
phase.  Polar compounds elute first, and nonpolar compounds elute later.

## Supplying readMsExperiment with metadata
Now that we've taken a first look at the data, I'm going to go back and supply
the `readMsExperiment` function with the relevant sampleData information so we
can use it downstream for peak grouping and statistical analyses. The original
MsExperiment objects will be replaced so they don't take up more memory.

Conveniently, we can build out sampleData from the assay data from
MetaboLights.  `dplyr` can filter the `study.info` table to split it into the
positive and negative ion mode information, and `stringr` can extract the year
value as a separate longitudinal variable. Sometimes, another package will
override the `filter` function definition from `dplyr`, so it's good to scope it
to the right one. The `study.info` table also has columns with duplicate names
that need to be renamed or removed to work with `dplyr`.

```{r}
study.info.m = study.info
colnames(study.info.m) = make.unique(colnames(study.info.m))
study.info.m = study.info.m |>
  mutate(
    polarity = str_extract(`Sample Name`, 'POS|NEG'),
    year = str_extract(`Sample Name`, '\\_\\d+\\_') |> str_replace_all('\\_', '') |> as.numeric()
  )
pos.data = readMsExperiment(pos.files, 
  sampleData = study.info.m |> filter(polarity == 'POS'))
```

##  Peak picking
The next step is to get some starting parameters for peak picking. First, we'll
zoom in on the early and late parts to refine the retention time window on which
peak picking will operate. `xcmsVis` provides `ggplot` version of `xcms` plots,
and `plotly` enables interactions, although it's slower than generating static
plots, so I'm not going to incorporate it into my `knitr` output.

```{r, eval=FALSE}
gplot(chromatogram(pos.data)) |> ggplotly()
```

Zooming in, chromatographic peaks start eluting at 25 s, which is the dead time.
This is the time it takes for unretained sample to pass through the column and
reach the detector. It can be determined based on the column geometry and flow
rate. For example, in this study, the authors used a 2.1 &times; 100 mm column
with a 5 mm guard column (notice the discrepancy in the study.info and the
Chromatography section of the protocol info), but didn't specify the flow rate.
These columns are usually run at 0.1 -- 0.2 mL/min. Let's check using the
observed dead time.  Column volume can be estimated, depending on the porosity,
as 0.7 &times; &pi; r^2^ &times; L, in &mu;L, where r is the column radius (half
the column diameter noted above) and L is the column length. Note this ignores
any dead volume contributions from extra-column tubing.

```{r}
dead.time = 25
col.vol = 0.7 * pi * (2.1/2)^2 * 105
est.flow.rate = 60 * col.vol / dead.time
print(est.flow.rate)
```

I was a bit off. It looks like they used a flow rate of 0.500 -- 0.600 mL/min,
but that's not unheard of in ultra-high performance liquid chromatography (UHPLC).

At the other end of the chromatograms, it looks like a diverter valve was used
to redirect the stream when the gradient was done, and I don't see any artifacts,
so the retention window for peak picking will be `c(25, 480)`. Since the same
chromatographic method was used, this window should work for the negative mode
data too.

Before delving into peak picking, we need to set up colors for plotting. I use
`colorspace` to generate my palettes. For longitudinal studies, a sequential
palette is appropriate. Simple A versus B studies can use diverging palettes.
```{r}
year.levels = sort(unique(study.info.m$year))
year.colors = sequential_hcl(length(year.levels), 'Viridis')
names(year.colors) = sprintf('%02d', year.levels)

# QC (NA year) gets its own neutral color appended separately, rather than
# occupying a slot in the year gradient. 'NA' starts with a letter, so it
# always alphabetically sorts after any 2-digit zero-padded number.
year.colors['NA'] = '#7f7f7f'
```

Now to peak picking.

Manually looking at the plotly TIC, the biggest peak (between 207 and 223 s,
probably caffeine) had a peakwidth of about 16 seconds, so I'll round up to 20
for the max peakwidth. A smaller peak had a peakwidth of around 8 seconds, so
I'll round down to 5 for the min peakwidth.  The `ppm` parameter can usually be
estimated by the instrument type. For TOFs and Q-TOFs, I've had to set this
between 20 -- 80 ppm for good peak detection.  Orbitraps are much more accurate
on the _m/z_ domain, so 10 ppm should be sufficient.

The `snr` parameter won't be touched because it's shown to not be
particularly useful.[@kumler2023picky]

```{r}
cwp = CentWaveParam(
  ppm = 10,
  peakwidth = c(5, 20)
)

peaks.pos = findChromPeaks(pos.data |> filterRt(c(25, 480)), param = cwp)

dir.create('_rds', showWarnings = FALSE)
saveRDS(peaks.pos, '_rds/peaks_pos.rds')
```

Since the biggest peak in tea is probably caffeine in positive mode, we'll use
that to test how peak picking went. We can use `webchem` to look up the
monoisotopic mass on PubChem, then calculate the theoretical _m/z_ for the
[M+H]+ ion.

```{r}
caffeine.cid = webchem::get_cid('caffeine')
caffeine.mass = webchem::pc_prop(caffeine.cid |> pull(cid), 'MonoisotopicMass')
caffeine.mz = as.numeric(caffeine.mass$MonoisotopicMass) + 1.00728
caffeine.mzr = LCMStools::calcPPM(caffeine.mz, 10)
```

Peak coloring can get tricky because the number of peaks often is greater than
the number of chromatograms to draw, so there's some complicated subsetting
required. I'll break it down as best as I can below.
```{r}
# We want to color the samples by how many years they were stored.
sampleData(peaks.pos)$year
# There are 22 samples in this set, with 7 different "year" conditions,
# including NA for the QC's
year.colors
# This will build out a list of what color each sample will be based on year
caffeine.colors =
  year.colors[str_replace_na(sprintf('%02d',sampleData(peaks.pos)$year))]
caffeine.colors

# The peaks need to be filtered by the caffeine _m/z_, otherwise we'll be
# getting back many more peaks from different ions
chromPeaks(peaks.pos |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])))[,'sample']

# Those peaks need to be mapped by which sample they are from to the
# list of caffeine colors
caffeine.colors[chromPeaks(peaks.pos |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])))[,'sample']]

# For peak backgrounds, the hex color needs an alpha value appended to it
# for transparency so it's easier to see overlaid peaks
paste0(caffeine.colors[chromPeaks(peaks.pos |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])))[,'sample']], '10')

# Sometimes I like to make this alpha value dynamic based on the number of
# samples. Since there are 22, dividing by a total opacity (if everything's overlaid)
# by number of samples, rounded down, yields 4% opacity. Since this is only 1
# digit, it needs to be padded for the hex color string.
caffeine.peakBg = paste0(
  caffeine.colors[chromPeaks(peaks.pos |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])))[,'sample']], 
  sprintf('%02d', floor(90 / length(caffeine.colors))))
caffeine.peakBg

# Ultimately, for coloring chromatograms and peaks, we end up with two lists:
# one for the ion current line (caffeine.colors), and one for the integrated
# peak background (caffeine.peakBg)

peaks.pos |> 
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  chromatogram() |> 
  plot(col = caffeine.colors, 
    peakBg = caffeine.peakBg,
    main = 'Caffeine peak results with initial cwp parameters')
```

The big split in the peak down the middle shows we need to refine these
parameters more. The peakwidth minimum and maximum should be increased so
there's no peak splitting. We can also check the mass accuracy drift from
the theoretical caffeine _m/z_.

```{r}
knitr::kable(chromPeaks(peaks.pos |> filterMz(c(caffeine.mzr[1], caffeine.mzr[2]))) |>
  as.data.frame() |>
  mutate(
    peakwidth = rtmax - rtmin,
    mz.error = abs(mz - caffeine.mz),
    ppm.error = 10^6 * (mz.error / caffeine.mz)
  )) |> kable_styling(full_width=TRUE) |> scroll_box(width='100%')

set.seed(17)
peaks.pos |> 
  filterFile(sample(seq_len(nrow(sampleData(peaks.pos))), 9)) |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  filterRt(c(200, 250)) |> plot(type='XIC')
```

From the table, we can narrow down the ppm threshold to 6 and adjust the
peakwidth.

```{r}
cwp = CentWaveParam(
  ppm = 6,
  peakwidth = c(10, 30)
)

peaks.pos = findChromPeaks(pos.data |> filterRt(c(25, 480)), param = cwp)

knitr::kable(chromPeaks(peaks.pos |> filterMz(c(caffeine.mzr[1], caffeine.mzr[2]))) |>
  as.data.frame() |>
  mutate(
    peakwidth = rtmax - rtmin,
    mz.error = abs(mz - caffeine.mz),
    ppm.error = 10^6 * (mz.error / caffeine.mz)
  )) |> kable_styling(full_width=TRUE) |> scroll_box(width='100%')

set.seed(17)
peaks.pos |> 
  filterFile(sample(seq_len(nrow(sampleData(peaks.pos))), 9)) |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  filterRt(c(200, 250)) |> plot(type='XIC')

par(mfrow=c(1,1))
peaks.pos |> 
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  chromatogram() |> 
  plot(col = caffeine.colors, 
    peakBg = caffeine.peakBg)
```

This looks better, but the integration window looks problematic. Caffeine tends
to tail in positive ESI mode for reasons unknown, and that tail isn't captured
here. It's probably fine since it's consistent, but we can adjust `cwp` to fix
it by changing the default parameter `integrate = 1` to `integrate = 2`.

```{r}
cwp = CentWaveParam(
  ppm = 6,
  peakwidth = c(10, 30),
  integrate = 2
)

peaks.pos = findChromPeaks(pos.data |> filterRt(c(25, 480)), param = cwp)

knitr::kable(chromPeaks(peaks.pos |> filterMz(c(caffeine.mzr[1], caffeine.mzr[2]))) |>
  as.data.frame() |>
  mutate(
    peakwidth = rtmax - rtmin,
    mz.error = abs(mz - caffeine.mz),
    ppm.error = 10^6 * (mz.error / caffeine.mz)
  )) |> kable_styling(full_width=TRUE) |> scroll_box(width='100%')

set.seed(17)
peaks.pos |> 
  filterFile(sample(seq_len(nrow(sampleData(peaks.pos))), 9)) |>
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  filterRt(c(200, 250)) |> plot(type='XIC')

par(mfrow=c(1,1))
peaks.pos |> 
  filterMz(c(caffeine.mzr[1], caffeine.mzr[2])) |>
  chromatogram() |> 
  plot(col = caffeine.colors, 
    peakBg = caffeine.peakBg)
```

This looks worse because it's not inaegrating the tail consistently. I'll settle
on the previous parameters.

```{r}
cwp = CentWaveParam(
  ppm = 6,
  peakwidth = c(10, 30)
)

peaks.pos = findChromPeaks(pos.data |> filterRt(c(25, 480)), param = cwp)
```

Another good control for quality is to look at random peaks to see how those
were treated.

```{r}
sample.colors = year.colors[str_replace_na(sprintf('%02d',sort(sampleData(peaks.pos)$year, na.last = TRUE)))]
sample.colors

set.seed(17)
par(mfrow=c(3,3))
walk(sample(seq_len(nrow(chromPeaks(peaks.pos))), 9), ~ {

  peak.info = chromPeaks(peaks.pos)[.x,]
  test.colors = sample.colors[str_replace_na(sprintf('%02d', sampleData(peaks.pos)$year))]

  peak.color = test.colors[peak.info['sample']]

  peaks.pos |>
    filterFile(peak.info['sample']) |>
    filterRt(c(peak.info['rtmin'] - 30, peak.info['rtmax'] + 30)) |>
    filterMz(c(peak.info['mzmin'], peak.info['mzmax'])) |>
  chromatogram() |>
  plot(col = peak.color, peakBg = paste0(peak.color, '70'))
})
```

Most of these look reasonable compared to a perfect Gaussian peak. It's never
going to be perfect with automated analysis. In the Kumler article, there are
new metrics for peak quality that use a peak correlation metric and signal to
noise estimate to filter out peaks, but I'll save those for another time.

## Peak grouping/correspondence

Each chromPeak is specific to a single sample. They need to be grouped across
samples to compare. Tracking peaks across samples is called "correspondence" in
`xcms`.

Note, usually people perform retention time alignment before correspondence.
In my opinion, if you need to align retention times, your chromatography is
not robust enough and you're not generating good data. Sometimes this is the
case when sequences don't use at least a few equlibration injections at the
beginning. This particular study looks well-behaved.

```{r}
pdp = PeakDensityParam(
  sampleGroups = sampleData(peaks.pos)$year,
  minFraction = 0.5,
  bw = 5
)
caffeine.peaks = peaks.pos |> filterMz(c(caffeine.mzr[1], caffeine.mzr[2]))

plotChromPeakDensity(chromatogram(caffeine.peaks), 
  col = sample.colors, 
  peakBg = paste0(sample.colors[chromPeaks(caffeine.peaks)[,'sample']], '05'), 
  param =  pdp)
```

That example looks pretty good so we'll apply the same parameters to the full
dataset.

```{r}
nrow(chromPeaks(peaks.pos))
peaks.pos.grouped = groupChromPeaks(peaks.pos, param = pdp)
nrow(featureDefinitions(peaks.pos.grouped))
```

We've now condensed the data from over 100,000 peaks to 5780 features, about
18-fold, which makes sense since there are 22 samples. Let's plot some
random features to see how well it worked.

```{r}
set.seed(17)
rand.features = featureChromatograms(
  peaks.pos.grouped, 
  features = sample(seq_len(nrow(featureDefinitions(peaks.pos.grouped))), 9)
)

knitr::kable(featureDefinitions(rand.features)) |>
  kable_styling(full_width=TRUE) |> scroll_box(width='100%')
knitr::kable(featureValues(rand.features)) |>
  kable_styling(full_width=TRUE) |> scroll_box(width='100%')

par(mfrow=c(3,3))
plot(rand.features, col = sample.colors,
  peakBg = paste0(sample.colors[chromPeaks(rand.features)[,'sample']], '10'))
```

`xcms` might not have integrated some features, so we have to fill in missing
peaks with "gap filling."

```{r}
peaks.pos.grouped.filled = fillChromPeaks(peaks.pos.grouped, param = ChromPeakAreaParam())

rand.features.filled = featureChromatograms(
  peaks.pos.grouped.filled,
  features = rownames(featureDefinitions(rand.features))
)
plot(rand.features.filled,
  col = sample.colors,
  peakBg = paste0(sample.colors[chromPeaks(rand.features.filled)[,'sample']], '10'))
saveRDS(peaks.pos.grouped.filled, '_rds/peaks_pos_grouped_filled.rds')
```

From the darker peak backgrounds, we can see that more peaks were integrated
with this method. This helps prevent zero'd integrations from skewing the
downstream analysis.

## Using the final results

We now basically have a table of feature intensities grouped across samples,
which proxies for metabolite abundances. We can use this for downstream 
statistical analyses to find what are the important features that change with
storage time. This is a lot less work than trying to identify every single peak
and then finding out most of them weren't important.

I usually start with a principal component analysis (PCA) to get an overview
of the data.

```{r}
pos.results = quantify(peaks.pos.grouped.filled, value = 'into', method = 'sum')
# ropls's plot.opls() does as.vector(parAsColFcVn) then re-derives its own
# factor internally, discarding any custom level order we set. So instead of
# pre-building a factor, we pass a plain character vector whose labels are
# zero-padded to match names(year.colors) exactly (see above) - that way
# ropls's own alphabetical re-sort lands on the same order as year.colors.
year.groups = ifelse(is.na(colData(pos.results)$year), 'NA',
  sprintf('%02d', colData(pos.results)$year))

pca.pos = opls(assay(pos.results, 'raw') |> t(), fig.pdfC='none')
plot(pca.pos, parPaletteVc = year.colors, parAsColFcVn = year.groups)
```

Nice. Each year condition clusters together, with the QC samples (prepared by
mixing equal amounts of each sample) in the middle.  The first principal
component explains 27% of the variance, with two features, FT4735 and FT5462,
heavily loaded on opposite sides of the p1 loading plot.  On the score plot, we
see short-term storage samples on the right and long-term storage samples on the
left, suggesting a linear relationship between storage time and metabolite
abundances.

We can go back to the feature definitions table to identify the two
standout features using `rowData()` from `SummarizedExperiment`.

```{r}
rowData(pos.results)[c('FT4735', 'FT5462'),]
big.feature.chromatograms = featureChromatograms(peaks.pos.grouped.filled, features = c('FT4735', 'FT5462'))

plot(big.feature.chromatograms,
  col = sample.colors,
  peakBg = paste0(sample.colors[chromPeaks(big.feature.chromatograms)[,'sample']], '10'))
chromPeaks(big.feature.chromatograms)[,'sample']
```

These features look pretty noisy and probably shouldn't be loaded so heavily.
Let's try a different scaling method, pareto scaling, which is often used in
biological data.

```{r}
pca.pos.pareto = opls(assay(pos.results, 'raw') |> t(), scaleC='pareto', fig.pdfC='none')
plot(pca.pos.pareto, parPaletteVc = year.colors, parAsColFcVn = year.groups)
```

You can see that scaling has a big effect on the results. The first principal
component now explains 41% of the variance, and we get more standout features.
Let's see how those look.

```{r}
important.features = getLoadingMN(pca.pos.pareto) |> as.data.frame() |>
  arrange(desc(abs(p1))) |>
  dplyr::slice(1:9) |>
  rownames()
important.features

important.feature.chromatograms = featureChromatograms(peaks.pos.grouped.filled,
  features = important.features)

plot(important.feature.chromatograms,
  col = sample.colors,
  peakBg = paste0(sample.colors[chromPeaks(important.feature.chromatograms)[,'sample']], '10'))
```

These look much more reasonable as important peaks, and we can clearly see
differences in metabolite abundances when coloring by storage time. Some may be
more noisy but some degree of false positives must be expected.

Notice that some of the features are actually redundant, or degenerate, meaning
they originate from the same compound or metabolite. In particular, the top left
and middle feature show the distinct peak shape of isomers coeluting, and they
elute at the same retention time. The middle right and bottom middle features also
overlap. Another way to reduce the dataset is to group these degenerate features
into compounds through the `CAMERA` package, which groups coeluting features into
"pseudospectra." However, this package hasn't been updated to work with the
new `xcms` interfaces introduced with version 3.0, so I haven't touched it in a while.
Another, more recent, package is `MsFeatures` by Johannes Rainer. This one
looks promising, and since it's so new I haven't gotten to implement it yet.
But the ideas behind it sound strong, as it groups features by how well they
correlate in the retention time dimension.

I digress. My point is that a feature is not a compound, but once we get
from features to compounds, data size can be reduced immensely.

Since this is a longitudinal study, we can use `year` as the independent variable
and feature intensities as dependent variables. We can use a linear model to see
how the important features change with storage time.

```{r}
lm.results = map_dfr(important.features, ~ {
  feature.intensities = assay(pos.results, 'raw')[.x,]
  lm.fit = lm(feature.intensities ~ colData(pos.results)$year)
  broom::tidy(lm.fit) |>
    mutate(feature = .x) |>
    mutate(
      adj.r.squared = broom::glance(lm.fit)$adj.r.squared
    )
})

lm.results

important.features.high.r2 = lm.results |> filter(adj.r.squared > 0.7) |> pull(feature) |> unique()
```

This narrowed it down to two features with an adjusted R-squared greater than
0.7, FT4766 and FT0396, or the plots corresponding to the 607.2477_411 _m/z_-RT
and 130.0507_69 _m/z_-RT features above.

Let's see how they quantitatively changed with storage time.

```{r}
map(important.features.high.r2, ~ {
  feature.intensities = assay(pos.results, 'raw')[.x,]

  plot.data = tibble(
    feature = .x,
    year = colData(pos.results)$year,
    intensity = feature.intensities
  )

  ggplot(plot.data, aes(x = year, y = intensity)) +
    geom_point() +
    geom_boxplot(aes(group = factor(year))) +
    geom_smooth(method = 'lm', color = 'red') +
    labs(title = paste0('Feature ', .x, ' vs. Storage Time'),
      x = 'Storage Time (years)',
      y = 'Feature Intensity') +
    theme_classic(base_size = 16)
})
```

So now we can say something smart about how storage time affects the 
metabolite abundances in white tea. Feature FT4766 decreases with storage time,
while FT0396 increases, with p-values of 1.23e-6 and 8.50e-6, respectively.

Quantitatively, FT0396 just about doubled over 11 years, and FT4766 decreased
to about 20% of its original abundance. This is a pretty big change, and it would be
interesting to see what these compounds are. We can try to identify them by
looking up their _m/z_ values in metabolite databases.

Looking up the _m/z_ from FT4766 across Phytohub, Massbank of Europe, and
Massbank of North America, only Phytohub gave a potential match for ligstroside
via an [M+2MeCN+H]+ adduct, while the other two yielded no results. We can check
this to see if the protonated ion, [M+H]+, coincides with that feature.

```{r}
ligstroside.mzr = LCMStools::calcPPM(524.1894+1.00728, 100)
featureDefinitions(peaks.pos.grouped.filled |> 
  filterMz(c(ligstroside.mzr[1], ligstroside.mzr[2])))
```

The resulting two filtered features don't match the retention time
of the ion, so this is likely something not in the databases.

For FT0396, several matches were found in Phytohub as [M+2H] adducts. We can
check if this makes sense by looking for an isotope peak at + 0.5 _m/z_ units
away from 130.0507, corresponding to where the 1C13 isotope would be if doubly
protonated.

```{r}
c13.mass.diff = 13.00335 - 12
c13.mass.diff
ft0396.mzr = LCMStools::calcPPM(
  featureDefinitions(peaks.pos.grouped.filled)['FT0396', 'mzmed'] + c13.mass.diff/2, 100
)

featureDefinitions(peaks.pos.grouped.filled |>
  filterMz(c(ft0396.mzr[1], ft0396.mzr[2])))
```

This also yielded no features, so it's likely a mismatch. Metabolite
identification is a difficult problem, and these two compounds would likely need
some structural information determined from MS/MS, or to be isolated with
multidimensional chromatography or fractionating sample preparation, then
analyzed by NMR to confirm their structures.

There are many other ways to analyze this data, and that may be a topic for
a future post. For now, this is a good starting point for exploring longitudinal
metabolomics data.

## Session Info
```{r}
sessionInfo()
```

## References

_Raw LC-MS data used in this post is from the MetaboLights study MTBLS13336 Shu
Ya Wei and Long Wen Li, available under [CC0 1.0
Universal](https://creativecommons.org/publicdomain/zero/1.0/) at 
https://www.ebi.ac.uk/metabolights/MTBLS13336. MetaboLights: Yurekten et al.,
2023._