---
title: "Exploratory metabolomic/longitudinal analysis of white tea"
author: "Josh Morimoto"
date: "July 11, 2026"
output: html_fragment
---

```{r, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

# Exploratory metabolomic/longitudinal analysis of white tea
Tea is an important subject for me because it was the model crop on which my
PhD thesis was based. At Tufts, we harvested tea grown in different seasons and
elevations to see how its environment changed its flavor and health properties,
all driven by its metabolism. Like many plants, tea uses chemical signals to
communicate with its environment, since it can't get up and walk away like an
animal or even a microbe could. Evolution has driven myriad metabolic pathways
to allow plants to defend themselves. This makes them particularly complex and
challenging subjects of research, slowing studies on how crop qualities will
change with the increasing threat of climate change.

For this demo/tutorial, I looked for data collected from _Camellia sinensis_
samples on MetaboLights and found one study in progress by Wei and Li
(https://www.ebi.ac.uk/metabolights/editor/MTBLS13336/overview).  This study
aimed to investigate the sensory and chemical metabolites, both endogenous to
the tea and exogenous from microbial contributions, over long-term storage. We
will focus on the metabolomics data herein.

## IDE considerations
I use VSCode with R and arf as the R frontend. Many people use RStudio or Python
with marimo or Jupyter notebooks but this is what works best for me. I also
manage R/Python environments with conda. For more info, follow Renkun Ken's blog
on writing R in VScode:
https://renkun.me/2019/12/11/writing-r-in-vscode-a-fresh-start/. He's my hero.

You'll also notice I don't really stick to the R style guidelines. I like my
`=` signs more than typing out `<-` everytime because I learned on Python.
It has never not worked for me. Conversely, I also like naming variables with
`.` rather than snake_case or camelCase. It looks better to my eyeballs.

## Libraries
```{r}
library(Spectra)
library(MsBackendMetaboLights)
library(purrr)
library(stringr)
library(dplyr)
library(curl)
library(LCMStools)
library(MsExperiment)
library(BiocParallel)

if(.Platform$OS == 'Windows'){
  register(bpstart(SnowParam(6, progressbar=FALSE)))
} else if(.Platform$OS == 'unix'){
  register(bpstart(MulticoreParam(6, progressbar=FALSE)))
}
```

For `httpgd` with VSCode and arf

```{r, eval=FALSE}
library(httpgd)
hgd()
hgd_browse()
```

## Loading the study data
First we load the study information from the MetaboLights repository using the
MTBLS ID. This is my first time personally using MetaboLights, so I'm getting a
view on how the data is organized.

```{r}
mtbls.repo = 'MTBLS13336'
all.files = mtbls_list_files(mtbls.repo)
all.files
```

Raw data is stored under *FILES* and study metadata in various formats in the
root directory.

Below is the assay information table, also accessible online at
https://www.ebi.ac.uk/metabolights/editor/MTBLS13336/assays.  This contains
sample names, data file names, and some experimental parameters.  If the data
were converted, we would also see a different file under the *Derived Spectral
Data File* column, but the authors kept them in Thermo's proprietary .raw
format. Otherwise, using the `Spectra` package and the `MsBackendMetaboLights`
backend, we could have loaded the *.mzML* files directly for analysis.

```{r}
study.info = read.table(paste0(mtbls_ftp_path(mtbls.repo), grep('^a_', all.files, value = TRUE)),
    sep = '\t', header = TRUE, check.names = FALSE)
study.info |> head()
study.info[, c('Raw Spectral Data File', 'Derived Spectral Data File')] |> head()
#specs = Spectra::Spectra('MTBLS13336', filePattern = '.raw', source = MsBackendMetaboLights())
```

Instead, we'll download the raw data and convert herein. This will take a while,
as high resolution mass spec data can be several gigabytes per file, so if
you're playing along at home, you can execute this block and go get a coffee.
In this case, they're ~80 MB per file. I've never worked with Orbitrap data so
that's pretty neat compared to Q-TOF profile data.

Note that I had to replace the `FILES/` prefix from the repo because I don't
need that in my local copy.

I thought parallelizing this with walk would make it go fast but it does not.
Probably faster to download a zip and unzip it locally.

```{r, eval=FALSE, progress=FALSE}
paste0(mtbls_ftp_path(mtbls.repo), mtbls.repo, '.zip')
dir.create('_raw', showWarnings=FALSE)
download.file(
  url = paste0(mtbls_ftp_path(mtbls.repo), mtbls.repo, '.zip'),
  destfile = paste0(mtbls.repo, '.zip')
)
```

That didn't work. MetaboLights doesn't zip raw data in this repo.
The `curl` package has a function `multi_download` that might work faster.

```{r, eval=FALSE, progress=FALSE}
dir.create('_raw', showWarnings=FALSE)

response.table = curl::multi_download(
  urls = paste0(
    mtbls_ftp_path(mtbls.repo), 
    study.info[, 'Raw Spectral Data File']), 
  destfiles = paste0(
    './_raw/',
    stringr::str_replace(study.info[, 'Raw Spectral Data File'], 'FILES/', '')),
  resume = TRUE
)

response.table[,'error'] |> head()
response.table$destfile |> head()
dir('_raw') |> head()
```

This didn't work either due to timeouts with the ftp server. In the end,
downloading the files sequentially with `download_file` was most reliable.

```{r}
dir.create('_raw', showWarnings=FALSE)
if(length(study.info[, 'Raw Spectral Data File']) != length(dir('_raw'))){
  walk(study.info[, 'Raw Spectral Data File'], ~ download.file(
    url = paste0(mtbls_ftp_path(mtbls.repo), .x),
    destfile = paste0('_raw/', stringr::str_replace(.x, 'FILES/', ''))),
    method = 'libcurl'
  )
}
```

## Data conversion
To read files with `xcms`, data need to be converted to *.mzML*
format with `MSConvert` from the `ProteoWizard` package. Since I'm on Linux,
I need a docker container to run this with a Windows image. On Windows, it can
run natively. The code below detects the host environment and chooses the appropriate
conversion command. The `convertDtoMZML` function from my `LCMStools` package
only works for Agilent *.d* files. Docker will prompt for a password for sudo
access. By using the Docker image, you agree to certain vendor licenses for
converting proprietary formats to *.mzML*.

```{r}
raw.files = dir('_raw', pattern = '\\.raw$', full.names = TRUE)

# Check if length mzML files matches length raw files
if(length(raw.files) > length(dir('_mzML', pattern = '\\.mzML$'))){
  if(.Platform$OS == 'Windows'){
    d.files |> walk(convertDtoMZML, path.out = '_mzML')
  } else if(.Platform$OS == 'unix'){
    walk(raw.files, ~ system(paste0(
      'sudo docker run -it --rm -v $PWD:/data proteowizard/pwiz-skyline-i-agree-to-the-vendor-licenses:latest wine msconvert /data/_raw/', basename(.), ' --filter "peakPicking true 1-" -o /data/_mzML --mzML'
      )
    ))
  }
}

mzml.files = dir('_mzML', pattern = '\\.mzML$', full.names = TRUE)
```

## Initial data inspection
First we'll plot the positive and negative TICs to make sure there aren't any
misbehaving data files or sample runs.

```{r}
mzml.files |> head()
pos.files = mzml.files[str_detect(mzml.files, 'POS')]
pos.files
neg.files = mzml.files[str_detect(mzml.files, 'NEG')]
pos.data = readMsExperiment(pos.files)
neg.data = readMsExperiment(neg.files)

plot(chromatogram(pos.data), main = 'Positive TIC')

plot(chromatogram(neg.data), main = 'Negative TIC')
```

It looks like nothing disastrous happened in the data collection. From the TICs,
it's hard to tell right away if there are any difference since the noise buries
most of the valuable data. Sometimes you can get away with focusing on a certain
region that's obviously different.

Looking closer, the polar region before 100 seconds in retention changes with
intensity in both the positive and negative ionization modes. In the next part,
we'll color code these TICs and actually do some peak picking on the _m/z_
level.

## Stay tuned for Part 2.
<!--
## Supplying readMsExperiment with metadata
Now that we've taken a first look at the data, I'm going to go back and supply
the `readMsExperiment` function with the relevant sampleData information so we
can use it downstream for peak grouping and statistical analyses. The original
MsExperiment objects will be replaced so they don't take up more memory.

Coneveniently, we can build out sampleData from the assay data from MetaboLights.
```{r}

```
-->
_Raw LC-MS data used in this post is from the MetaboLights study MTBLS13336 Shu
Ya Wei and Long Wen Li, available under [CC0 1.0
Universal](https://creativecommons.org/publicdomain/zero/1.0/) at 
https://www.ebi.ac.uk/metabolights/MTBLS13336. MetaboLights: Yurekten et al.,
2023._